---
date: "2025-11-10"
output: html_document
---

# Unbiased Relative Difference Analysis of the Interim and Legacy Products

> Update: November 2025

## Objectives

Calculate the mean unbiased relative difference $\psi^{\text{netPP}}$ of the interim VIIRS netPP and legacy MODIS netPP products for each month from the timeseries shared between the two sensors.

We are using this statistic to track the similarities in netPP values between interim VIIRS and legacy MODIS datasets to provide validation that the interim netPP products can be reliably used for continuity in long-term productivity analyses.

For **VIIRS-SNPP**, the result will be a 120-month (10-year) mean monthly timeseries of $\psi^{\text{netPP}}$.

For **VIIRS-NOAA20**, the result will be a 60-month (5-year) mean monthly timeseries of $\psi^{\text{netPP}}$.

The unbiased relative difference $\psi^{\text{netPP}}$ is calculated for each pixel as follows:

$$
\psi^{\text{netPP}} = \frac{\text{netPP}_{\text{VIIRS}} - \text{netPP}_{\text{MODIS}}}{\left( \text{netPP}_{\text{VIIRS}} + \text{netPP}_{\text{MODIS}} \right) \div 2}
$$ Where:

-   The difference between VIIRS-based netPP ($\text{netPP}_{\text{VIIRS}}$) and MODIS-based netPP ($\text{netPP}_{\text{MODIS}}$) values for that pixel is divided by the mean of $\text{netPP}_{\text{VIIRS}}$ and $\text{netPP}_{\text{MODIS}}$.

Normalizing by the mean of the two values avoids arbitrarily selecting one product as a reference.

------------------------------------------------------------------------

## Datasets Overview

We will be creating the pixel-by-pixel unbiased relative difference ($\psi^{\text{netPP}}$) for each month at a 9km resolution across the VIIRS-NOAA20 and MODIS-Aqua datasets. Here are the available datasets including VIIRS-SNPP:

1.  **Primary Productivity, MODIS Aqua, Science Quality, Global, 9km, 2013-2022, (Monthly Composite)**

-   Distributed via the West Coast Node ERDDAP dataset at the following link: <https://coastwatch.pfeg.noaa.gov/wcn/erddap/griddap/productivity_modis_aqua_monthly_9km.graph>

2.  **Primary Productivity, VIIRS SNPP, Science Quality, Global, 9km, 2013-2022 (Monthly Composite)**

-   Distributed via the West Coast Node ERDDAP dataset at the following link: <https://coastwatch.pfeg.noaa.gov/wcn/erddap/griddap/productivity_viirs_snpp_monthly_9km.graph>

3.  **Primary Productivity, VIIRS NOAA20, Science Quality, Global, 9km, 2018-2022, (Monthly Composite)**

-   Distributed via the West Coast Node ERDDAP dataset at the following link: <https://coastwatch.pfeg.noaa.gov/wcn/erddap/griddap/productivity_viirs_noaa20_monthly_9km.graph>

## Resource requirements

-   **R Studio** with the modules included within the *Install and Load Required Packages* section below

-   **Internet connection**

## Tutorial for this notebook

For this tutorial, we will be using the **VIIRS-NOAA20** and **MODIS-Aqua** datasets to calculate the mean unbiased relative difference from 2018-2022.

## Install and Load Required Packages

```{r}
pkges = installed.packages()[,"Package"]
# Function to check if pkgs are installed, install missing pkgs, and load
pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE,repos='http://cran.us.r-project.org')
    if(!require(x,character.only = TRUE)) stop(x, " :Package not found")
  }
}

# create list of required packages
list.of.packages <- c("ncdf4", "rerddap","plotdap", "parsedate", 
                      "sp", "ggplot2", "RColorBrewer", "sf", 
                      "reshape2", "maps", "mapdata", 
                      "jsonlite", "rerddapXtracto", "dplyr",
                      "lubridate", "rlang")

# Run install and load function
for (pk in list.of.packages) {
  pkgTest(pk)
}

# create list of installed packages
pkges = installed.packages()[,"Package"]
```

## Create Global Variables

Global variables are used to set up the directory paths required for processing trends.

-   **BASE_DIR** is the base directory defined for the project, and the root for all other subdirectories.
-   **ODATA_DIR** is where the output psi file will be located.

```{r}
ROOT_DIR <- "/Users/madisonrichardson/netpp"
ODATA_DIR <- file.path(ROOT_DIR, "data", "matrix", "psi")

dir_list <- c(ROOT_DIR, ODATA_DIR)

# Create if missing
invisible(lapply(dir_list, dir.create, showWarnings = FALSE, recursive = TRUE))

cat(length(dir_list), "directories validated\n")
```

## Make Some Useful Functions

```{r}

#' Center timestamps on the 16th (and optionally at midnight)
#'
#' @description
#' For monthly composites, standardize timestamps so they fall on the **16th**
#' of each month. Designed for data frames returned by `rerddap::griddap(...)$data`.
#' Optionally zeroes the time-of-day to `00:00:00` (recommended for joins).
#'
#' @param x Either:
#'   - a `data.frame`/`tibble` with a date-time column, or
#'   - a `POSIXct`/`Date` vector.
#' @param time_col Character. Name of the time column if `x` is a data frame.
#'   Ignored otherwise. Default `"time"`.
#' @param normalize_midnight Logical. If `TRUE`, set hour/min/sec to `00:00:00`.
#'   Default `TRUE` so datasets with different hours (e.g., 08Z vs 12Z) align.
#' @param quiet Logical. If `FALSE`, prints a short message. Default `TRUE`.
#'
#' @returns
#' Object of the same class as `x`, with timestamps on the 16th (and, by default,
#' at midnight). If input is a `Date`, it is returned as `Date`; if `POSIXct`,
#' returned as `POSIXct`. For data frames, only `time_col` is modified.
#'

ensure_time_centered_on_16th <- function(x,
                                         time_col = "time",
                                         normalize_midnight = TRUE,
                                         quiet = TRUE) {
  # Vector method --------------------------------------------------------------
  if (inherits(x, c("POSIXct", "POSIXlt", "Date"))) {
    # Track type to return in same class
    is_date <- inherits(x, "Date")
    tz_attr <- if (!is_date) attr(x, "tzone") else "UTC"
    
    # Work in POSIXlt for component edits
    lt <- if (is_date) as.POSIXlt(x) else as.POSIXlt(x, tz = tz_attr)
    all16 <- all(as.integer(format(lt, "%d")) == 16L, na.rm = TRUE)
    
    if (!all16 && !quiet) message("Replacing day with 16...")
    lt$mday <- 16L
    
    if (normalize_midnight) {
      lt$hour <- 0L; lt$min <- 0L; lt$sec <- 0
    }
    
    if (is_date) {
      return(as.Date(lt))
    } else {
      return(as.POSIXct(lt, tz = tz_attr))
    }
  }
  
  # Data-frame method ----------------------------------------------------------
  if (is.data.frame(x)) {
    if (!time_col %in% names(x)) stop("Column '", time_col, "' not found.")
    t <- x[[time_col]]
    
    if (!inherits(t, c("POSIXct", "POSIXlt", "Date"))) {
      stop("Column '", time_col, "' must be POSIXct/POSIXlt or Date.")
    }
    
    # Preserve original class for the column
    is_date <- inherits(t, "Date")
    tz_attr <- if (!is_date) attr(t, "tzone") else "UTC"
    
    lt <- if (is_date) as.POSIXlt(t) else as.POSIXlt(t, tz = tz_attr)
    all16 <- all(as.integer(format(lt, "%d")) == 16L, na.rm = TRUE)
    
    if (!all16 && !quiet) message("Replacing day with 16...")
    lt$mday <- 16L
    
    if (normalize_midnight) {
      lt$hour <- 0L; lt$min <- 0L; lt$sec <- 0
    }
    
    x[[time_col]] <- if (is_date) as.Date(lt) else as.POSIXct(lt, tz = tz_attr)
    return(x)
  }
  
  stop("Unsupported input: pass a POSIXct/Date vector or a data.frame with '", time_col, "'.")
}

```

```{r}
#' Unbiased relative difference (ψ) between two numeric datasets
#'
#' @title Calculate ψ = (A - B) / ((A + B)/2) with NA masking
#'
#' @description
#' Computes the **unbiased relative difference** (ψ) between a *minuend* dataset
#' `A` and a *subtrahend* dataset `B`, defined as
#' \deqn{\psi = \frac{A - B}{(A + B)/2}.}
#' Missing values are respected via `NA`; wherever either input is `NA`, the
#' result is `NA`. Division-by-zero (i.e., where the mean denominator is exactly
#' zero) is handled according to `zero_denominator`.
#'
#' @details
#' - *Minuend* is the quantity **from which** another is subtracted (A).
#' - *Subtrahend* is the quantity **that is subtracted** (B).
#' - Inputs must be numeric vectors, matrices, or arrays of identical shape.
#' - If you are starting from masked arrays (Python `numpy.ma` semantics),
#'   convert masked elements to `NA_real_` before calling this function.
#' - For gridded data classes that support vectorized arithmetic (e.g., **stars**,
#'   **terra**), you can usually pass them directly if they share geometry and
#'   alignment; otherwise extract numeric arrays first and then re-wrap.
#'
#' @param minuend_data Numeric vector/matrix/array. The dataset **A** (minuend).
#' @param subtrahend_data Numeric vector/matrix/array. The dataset **B** (subtrahend).
#'   Must be the same shape as `minuend_data`.
#' @param zero_denominator Character scalar controlling how to treat elements
#'   where the average \eqn{(A+B)/2} equals exactly 0. One of:
#'   - `"NA"` (default): set ψ to `NA_real_` at those locations.
#'   - `"Inf"`: keep machine `Inf/-Inf` results.
#'   - `"Zero"`: force ψ to `0` at those locations.
#'
#' @returns
#' A numeric object of the same type and dimensions as the inputs (vector/matrix/array)
#' containing ψ, with `NA` wherever inputs were `NA` (or where division-by-zero was
#' mapped to `NA`).
#'

calculate_psi_dr <- function(minuend_data,
                             subtrahend_data,
                             zero_denominator = c("NA", "Inf", "Zero")) {
  zero_denominator <- match.arg(zero_denominator)
  
  if (!is.numeric(minuend_data) || !is.numeric(subtrahend_data)) {
    stop("Both 'minuend_data' and 'subtrahend_data' must be numeric.")
  }
  if (!identical(dim(minuend_data), dim(subtrahend_data))) {
    # For vectors, dim() is NULL; identical(NULL, NULL) is TRUE.
    if (!(is.null(dim(minuend_data)) && is.null(dim(subtrahend_data)) &&
          length(minuend_data) == length(subtrahend_data))) {
      stop("Inputs must have identical shapes (same length/dimensions).")
    }
  }
  
  # Element-wise average (denominator) and numerator
  avg <- (minuend_data + subtrahend_data) / 2
  num <- (minuend_data - subtrahend_data)
  
  # Compute psi with standard vectorized division
  psi <- num / avg
  
  # Handle exact-zero denominators in a controlled way.
  # Note: comparisons to zero are safe here because 'avg' is computed
  # from finite arithmetic; NA positions are excluded.
  denom_zero_idx <- !is.na(avg) & (avg == 0)
  
  if (any(denom_zero_idx, na.rm = TRUE)) {
    if (zero_denominator == "NA") {
      psi[denom_zero_idx] <- NA_real_
    } else if (zero_denominator == "Zero") {
      psi[denom_zero_idx] <- 0
    } else {
      # "Inf": leave as produced by num/avg (Inf/-Inf or NaN if num also 0)
      # No action needed.
      invisible(NULL)
    }
  }
  
  # Ensure that if either input was NA, output is NA (already true with base ops).
  # This line is defensive in case future modifications alter NA propagation:
  na_mask <- is.na(minuend_data) | is.na(subtrahend_data)
  if (any(na_mask)) psi[na_mask] <- NA_real_
  
  return(psi)
}

```

```{r}

#' Write ψ (unbiased relative difference) to an ERDDAP-style NetCDF file
#'
#' @description
#' Create a CF/ERDDAP-style NetCDF file containing a 3-D field of unbiased
#' relative difference (ψ) on a regular lon/lat/time grid. The function:
#' \enumerate{
#'   \item Validates the input array dimensions against the coordinate vectors.
#'   \item Derives spatial bounds from the latitude/longitude vectors.
#'   \item Constructs an output filename that encodes sensor names and
#'         lat/lon bounds.
#'   \item Defines CF-style dimensions (`time`, `latitude`, `longitude`) and a
#'         `psi` variable on \code{[lon, lat, time]}.
#'   \item Writes the data to disk and attaches ERDDAP-like coordinate and
#'         global metadata (patterned after the psi CDL template).
#' }
#'
#' @param psi_array Numeric 3-D array of ψ values with dimensions
#'   \code{[time, lat, lon]}. This is usually the result of computing
#'   ψ pixel-wise between two gridded products and stacking over time.
#' @param time_secs Numeric vector of length \code{dim(psi_array)[1]} giving
#'   time coordinates in \strong{seconds since 1970-01-01T00:00:00Z}.
#' @param lat_vals Numeric vector of length \code{dim(psi_array)[2]} giving
#'   latitude coordinates in degrees north.
#' @param lon_vals Numeric vector of length \code{dim(psi_array)[3]} giving
#'   longitude coordinates in degrees east (e.g., \code{[-180, 180]} or
#'   \code{[0, 360]}).
#' @param sensor1 Character scalar giving the name of the VIIRS (or other
#'   interim) sensor, e.g. \code{"noaa20"} or \code{"snpp"}. Used in the
#'   filename and title as the minuend dataset in ψ = VIIRS – MODIS.
#' @param sensor2 Character scalar giving the name of the legacy/reference
#'   sensor, default \code{"modis"}. Used in the filename and title as the
#'   subtrahend dataset in ψ = sensor1 – sensor2.
#' @param out_dir Character path to the directory where the NetCDF file will
#'   be written. The directory will be created if it does not exist.
#' @param region_label Character label describing the spatial subset, used
#'   in the global \code{title} and \code{summary} attributes
#'   (e.g. \code{"West Coast subset"}).
#' @param overwrite Logical. If \code{FALSE} (default), the function stops
#'   with an error if a file with the same name already exists. If
#'   \code{TRUE}, an existing file is deleted and recreated.
#'
#' @details
#' The output filename has the form:
#' \preformatted{
#'   netpp_psi_<sensor1>_<sensor2>_lon<min>_to_<max>_lat<min>_to_<max>.nc
#' }
#' where the lat/lon bounds are taken from \code{lat_vals} and \code{lon_vals}
#' and formatted to one decimal place.
#'
#' Internally, the \code{psi_array} is transposed from
#' \code{[time, lat, lon]} to \code{[lon, lat, time]} before writing, so that
#' the resulting NetCDF is consistent with ERDDAP’s \code{[longitude, latitude,
#' time]} ordering. The \code{time} dimension is created as an unlimited
#' dimension to allow appending in other workflows.
#'
#' The NetCDF file includes:
#' \itemize{
#'   \item CF/ERDDAP-style attributes on \code{time}, \code{latitude},
#'         \code{longitude}, and \code{psi}, following the psi CDL template.
#'   \item Geospatial coverage metadata (\code{geospatial_*} and
#'         \code{southernmost_latitude} etc.).
#'   \item A project/title/summary block tailored to the interim
#'         primary productivity use case.
#' }
#'
#' @return
#' Invisibly returns the full file path (\code{character}) of the created
#' NetCDF file. A message is printed to the console indicating where the
#' file was written.
#'
#' @export
write_psi_netcdf <- function(psi_array,
                             time_secs,
                             lat_vals,
                             lon_vals,
                             sensor1,
                             sensor2 = "modis",
                             out_dir,
                             region_label = "West Coast subset",
                             overwrite = FALSE) {
  # Ensure ncdf4 is available
  if (!requireNamespace("ncdf4", quietly = TRUE)) {
    stop("Package 'ncdf4' is required but not installed.")
  }

  # Basic checks --------------------------------------------------------------
  if (!is.numeric(psi_array)) stop("'psi_array' must be numeric.")
  if (length(dim(psi_array)) != 3L) {
    stop("'psi_array' must be a 3-D array [time, lat, lon].")
  }

  n_time <- length(time_secs)
  n_lat  <- length(lat_vals)
  n_lon  <- length(lon_vals)

  if (!identical(dim(psi_array), c(n_time, n_lat, n_lon))) {
    stop(
      "dim(psi_array) must be [length(time_secs), length(lat_vals), length(lon_vals)].\n",
      "Found: ", paste(dim(psi_array), collapse = " x "), " vs ",
      n_time, " x ", n_lat, " x ", n_lon
    )
  }

  # Derive spatial bounds for filename & metadata -----------------------------
  lat_min <- min(lat_vals); lat_max <- max(lat_vals)
  lon_min <- min(lon_vals); lon_max <- max(lon_vals)

  # Compact numeric formatting for filename
  fmt <- function(x) sprintf("%.1f", x)

  nc_filename <- sprintf(
    "netpp_psi_%s_%s_lon%s_to_%s_lat%s_to_%s.nc",
    tolower(sensor1),
    tolower(sensor2),
    fmt(lon_min), fmt(lon_max),
    fmt(lat_min), fmt(lat_max)
  )

  if (!dir.exists(out_dir)) {
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  }
  nc_file_path <- file.path(out_dir, nc_filename)

  # Dimensions and variable definition ----------------------------------------
  dim_time <- ncdf4::ncdim_def(
    name  = "time",
    units = "seconds since 1970-01-01T00:00:00Z",
    vals  = time_secs,
    unlim = TRUE,
    create_dimvar = TRUE
  )

  dim_lat <- ncdf4::ncdim_def(
    name  = "latitude",
    units = "degrees_north",
    vals  = lat_vals,
    create_dimvar = TRUE
  )

  dim_lon <- ncdf4::ncdim_def(
    name  = "longitude",
    units = "degrees_east",
    vals  = lon_vals,
    create_dimvar = TRUE
  )

  # psi stored as [lon, lat, time] in NetCDF → dims in R order [lon, lat, time]
  var_psi <- ncdf4::ncvar_def(
    name     = "psi",
    units    = "1",  # unitless
    dim      = list(dim_lon, dim_lat, dim_time),
    longname = "Summation for Calculating Psi",
    missval  = -999.0,
    prec     = "float"
  )

  # ------------------------------------------------------------
  # Create NetCDF file and write data
  # ------------------------------------------------------------
  if (file.exists(nc_file_path)) {
    if (!overwrite) {
      stop(
        "NetCDF file already exists at:\n  ", nc_file_path,
        "\nSet overwrite = TRUE or delete the file manually if you want to recreate it."
      )
    } else {
      message("File exists, overwriting: ", nc_file_path)
      unlink(nc_file_path)
    }
  }

  nc <- ncdf4::nc_create(nc_file_path, vars = list(var_psi))

  # psi_array is [time, lat, lon]; NetCDF expects [lon, lat, time]
  psi_for_nc <- aperm(psi_array, perm = c(3, 2, 1))
  ncdf4::ncvar_put(nc, "psi", psi_for_nc)

  # Coordinate variables (dimvars auto-created by ncdim_def)
  ncdf4::ncvar_put(nc, "time",      time_secs)
  ncdf4::ncvar_put(nc, "latitude",  lat_vals)
  ncdf4::ncvar_put(nc, "longitude", lon_vals)

  # ------------------------------------------------------------
  # Add ERDDAP-style attributes (mirroring the psi CDL)
  # ------------------------------------------------------------
  now <- Sys.time()

  # Coordinate axis metadata: time
  ncdf4::ncatt_put(nc, "time", "_CoordinateAxisType", "Time",      prec = "text")
  ncdf4::ncatt_put(nc, "time", "axis",                "T",         prec = "text")
  ncdf4::ncatt_put(nc, "time", "calendar",            "gregorian", prec = "text")
  ncdf4::ncatt_put(nc, "time", "ioos_category",       "Time",      prec = "text")
  ncdf4::ncatt_put(nc, "time", "long_name",           "Time",      prec = "text")
  ncdf4::ncatt_put(nc, "time", "standard_name",       "time",      prec = "text")
  ncdf4::ncatt_put(nc, "time", "time_origin",
                   "01-JAN-1970 00:00:00",           prec = "text")

  # latitude
  ncdf4::ncatt_put(nc, "latitude", "_CoordinateAxisType", "Lat",       prec = "text")
  ncdf4::ncatt_put(nc, "latitude", "axis",                "Y",         prec = "text")
  ncdf4::ncatt_put(nc, "latitude", "ioos_category",       "Location",  prec = "text")
  ncdf4::ncatt_put(nc, "latitude", "long_name",           "Latitude",  prec = "text")
  ncdf4::ncatt_put(nc, "latitude", "standard_name",       "latitude",  prec = "text")
  ncdf4::ncatt_put(nc, "latitude", "valid_min",           lat_min,     prec = "float")
  ncdf4::ncatt_put(nc, "latitude", "valid_max",           lat_max,     prec = "float")

  # longitude
  ncdf4::ncatt_put(nc, "longitude", "_CoordinateAxisType", "Lon",       prec = "text")
  ncdf4::ncatt_put(nc, "longitude", "axis",                "X",         prec = "text")
  ncdf4::ncatt_put(nc, "longitude", "ioos_category",       "Location",  prec = "text")
  ncdf4::ncatt_put(nc, "longitude", "long_name",           "Longitude", prec = "text")
  ncdf4::ncatt_put(nc, "longitude", "source_name",         "cols",      prec = "text")
  ncdf4::ncatt_put(nc, "longitude", "standard_name",       "longitude", prec = "text")
  ncdf4::ncatt_put(nc, "longitude", "valid_min",           lon_min,     prec = "float")
  ncdf4::ncatt_put(nc, "longitude", "valid_max",           lon_max,     prec = "float")

  # psi var attrs
  ncdf4::ncatt_put(nc, "psi", "ioos_category", "Other",    prec = "text")
  ncdf4::ncatt_put(nc, "psi", "long_name",     "Summation for Calculating Psi",
                   prec = "text")
  ncdf4::ncatt_put(nc, "psi", "missing_value", -999.0,     prec = "float")
  ncdf4::ncatt_put(nc, "psi", "units",         "1",        prec = "text")

  # Global attributes ---------------------------------------------------------
  lat_step <- if (length(lat_vals) > 1L) abs(diff(lat_vals)[1]) else NA_real_
  lon_step <- if (length(lon_vals) > 1L) abs(diff(lon_vals)[1]) else NA_real_

  ncdf4::ncatt_put(nc, 0, "cdm_data_type",            "Grid",       prec = "text")
  ncdf4::ncatt_put(nc, 0, "map_projection",           "geographic", prec = "text")
  # From psi CDL: "PDM"
  ncdf4::ncatt_put(nc, 0, "time_coverage_resolution", "PDM",        prec = "text")

  ncdf4::ncatt_put(
    nc, 0, "project",
    "NOAA CoastWatch Interim Primary Productivity",
    prec = "text"
  )

  ncdf4::ncatt_put(
    nc, 0, "title",
    sprintf(
      "Pixel by pixel Psi, VIIRS %s minus %s Aqua",
      toupper(sensor1), toupper(sensor2)
    ),
    prec = "text"
  )

  ncdf4::ncatt_put(
    nc, 0, "summary",
    sprintf(
      paste0(
        "The unbiased relative difference (Psi) between primary productivity (netPP) ",
        "calculated using VIIRS %s data and %s Aqua data. For each pixel, psi is ",
        "calculated as netPP(%s) - netPP(%s) divided by the mean of netPP(%s) and netPP(%s) ",
        "as described by Mélina et al. 2017. Primary productivity was calculated as ",
        "described by Behrenfeld and Falkowski 1997. Input data for primary productivity ",
        "were obtained from NASA and included chlorophyll_a, sea surface temperature, ",
        "and photosynthetically active radiation from either MODIS Aqua or VIIRS %s."
      ),
      toupper(sensor1), toupper(sensor2),
      tolower(sensor1), tolower(sensor2),
      tolower(sensor1), tolower(sensor2),
      toupper(sensor1)
    ),
    prec = "text"
  )

  ncdf4::ncatt_put(
    nc, 0, "date_created",
    format(as.POSIXct(now, tz = "UTC"), "%Y-%m-%dT%H:%M:%SZ"),
    prec = "text"
  )

  ncdf4::ncatt_put(nc, 0, "latitude_units",        "degrees_north", prec = "text")
  ncdf4::ncatt_put(nc, 0, "longitude_units",       "degrees_east",  prec = "text")
  ncdf4::ncatt_put(nc, 0, "southernmost_latitude", lat_min,         prec = "float")
  ncdf4::ncatt_put(nc, 0, "northernmost_latitude", lat_max,         prec = "float")
  ncdf4::ncatt_put(nc, 0, "westernmost_longitude", lon_min,         prec = "float")
  ncdf4::ncatt_put(nc, 0, "easternmost_longitude", lon_max,         prec = "float")
  ncdf4::ncatt_put(nc, 0, "geospatial_lat_min",    lat_min,         prec = "float")
  ncdf4::ncatt_put(nc, 0, "geospatial_lat_max",    lat_max,         prec = "float")
  ncdf4::ncatt_put(nc, 0, "geospatial_lon_min",    lon_min,         prec = "float")
  ncdf4::ncatt_put(nc, 0, "geospatial_lon_max",    lon_max,         prec = "float")

  if (!is.na(lat_step)) {
    ncdf4::ncatt_put(nc, 0, "geospatial_lat_resolution", lat_step, prec = "float")
    ncdf4::ncatt_put(nc, 0, "latitude_step",             lat_step, prec = "float")
  }
  if (!is.na(lon_step)) {
    ncdf4::ncatt_put(nc, 0, "geospatial_lon_resolution", lon_step, prec = "float")
    ncdf4::ncatt_put(nc, 0, "longitude_step",            lon_step, prec = "float")
  }

  ncdf4::ncatt_put(nc, 0, "geospatial_lat_units",    "degrees_north", prec = "text")
  ncdf4::ncatt_put(nc, 0, "geospatial_lon_units",    "degrees_east",  prec = "text")

  # From psi CDL:
  ncdf4::ncatt_put(nc, 0, "spatialResolution", "9.28 km",           prec = "text")
  ncdf4::ncatt_put(
    nc, 0, "proj4_string",
    "+proj=eqc +lat_ts=0 +lat_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs +lon_0=0.000000",
    prec = "text"
  )

  ncdf4::ncatt_put(
    nc, 0, "creator_name",
    "NOAA CoastWatch West Coast Node", prec = "text"
  )
  ncdf4::ncatt_put(
    nc, 0, "creator_url",
    "https://coastwatch.pfeg.noaa.gov/", prec = "text"
  )

  ncdf4::nc_close(nc)

  message("Wrote ERDDAP-like psi NetCDF to: ", nc_file_path)

  invisible(nc_file_path)
}

```

## Set Up Parameters

Initialize the parameters required for the unbiased difference calculation:

-   **start_date** and **end_date**: Define the date range for the analysis (format: YYYY-MM-DDTXX:XX:XXZ). Monthly composites are centered on the 16th of each month.

-   **sensor1** and **sensor2**: Specify the two satellite sensors for unbiased relative difference analysis ('modis', 'snpp', or 'noaa20').

-   **ncvar**: The name of the variable being analyzed (we are using 'productivity').

-   **overwrite**: If 'True", existing output files will be replaced.

```{r}
start_date_noaa20 = "2018-01-16T12:00:00Z"
end_date_noaa20 = "2022-12-16T12:00:00Z"

start_date_modis = "2018-01-16T08:00:00Z"
end_date_modis = "2022-12-16T08:00:00Z"

# Define sensor (either "snpp", "noaa20", or "modis")
sensor1 = "noaa20"
sensor2 = "modis"

# Valid sensors list
valid_sensors <- c("modis", "snpp", "noaa20")

# Check both sensors are valid
for (s in c(sensor1, sensor2)) {
  if (!s %in% valid_sensors) {
    stop(sprintf("The sensor '%s' is not a valid option. Choose from: %s",
                 s, paste(valid_sensors, collapse = ", ")))
  }
}

cat("Sensors validated:", sensor1, "and", sensor2, "\n")

ncvar = "productivity"
overwrite = FALSE

```

## Select Satellite Dataset from ERDDAP

We will be using the VIIRS-NOAA20 - MODIS-AQUA netpp datasets from the West Coast Node ERDDAP Server. The dataset IDs are: **productivity_viirs_noaa20_monthly_9km** and **productivity_modis_aqua_monthly_9km**. We will use the info function from the **rerddap** package to first obtain information about the dataset of interest, then we will import the data.

```{r}
# Set ERDDAP URL
erddap_url = "https://coastwatch.pfeg.noaa.gov/wcn/erddap/"

# Set dataset IDs
noaa20_id = "productivity_viirs_noaa20_monthly_9km"
modis_id = "productivity_modis_aqua_monthly_9km"


# Get datasets info
noaa20_dataInfo <- rerddap::info(noaa20_id, url=erddap_url)  
print(noaa20_dataInfo)

modis_dataInfo <- rerddap::info(modis_id, url=erddap_url)  
print(modis_dataInfo)

```

## Subset Datasets

The datasets are too large to use the entire globe, so we are subsetting to use the West Coast of the US.

```{r}
# Define West Coast latitude and longitudes
wc_lon <- c(-135, -115)
wc_lat <- c(30, 50)

# Subset noaa20 dataset
noaa20_ds <- griddap(
  datasetx = noaa20_id,
  url = erddap_url,
  time = c(start_date_noaa20, end_date_noaa20),
  longitude = wc_lon,
  latitude = wc_lat,
  fields = ncvar
)

print(noaa20_ds)

# Subset modis dataset
modis_ds <- griddap(
  datasetx = modis_id,
  url = erddap_url,
  time = c(start_date_modis, end_date_modis),
  longitude = wc_lon,
  latitude = wc_lat,
  fields = ncvar
)

print(modis_ds)

```

## Convert griddap() output to data frames and standardize time

The `griddap()` function returns its results in a list structure where the actual data are stored in the `$data` element. Here, we extract those data tables for VIIRS-NOAA20 and MODIS-Aqua and convert the `time` column to `POSIXct` in UTC so the two products use a consistent time format before we align and compare them.

```{r}
noaa20_df <- noaa20_ds$data
modis_df <- modis_ds$data

# Make sure time is POSIXct before calling the function
noaa20_df$time <- as.POSIXct(noaa20_df$time, tz = "UTC")
modis_df$time <- as.POSIXct(modis_df$time, tz = "UTC")

```

## Normalize timestamps to monthly centers (16th of each month)

The MODIS and VIIRS monthly composites do not always share identical timestamp formats, some products use 08:00Z, others 12:00Z, and occasionally different days within the month. To ensure a clean one-to-one match between the two datasets, we standardize all timestamps so they fall on the **16th of each month at 00:00 UTC**, which is the convention used in ERDDAP monthly products.

After centering the timestamps, we create a `date_key` column that stores only the date (YYYY-MM-16). This key is used to merge MODIS and VIIRS pixels by their shared month.

```{r}
noaa20_df <- ensure_time_centered_on_16th(noaa20_df, time_col = "time")
modis_df <- ensure_time_centered_on_16th(modis_df, time_col = "time")

# Make a date key (now all should be YYYY-mm-16)
modis_df   <- modis_df   %>% mutate(date_key = as.Date(time))
noaa20_df <- noaa20_df %>% mutate(date_key = as.Date(time))

```

## Identify months with data from both sensors

Before computing ψ, we restrict the analysis to months where **both** MODIS-Aqua and VIIRS-NOAA20 have valid data. We do this by intersecting the `date_key` columns from each dataset (each key represents a monthly composite centered on the 16th). The resulting `common_dates` vector defines the shared time axis for all subsequent ψ calculations.

We also include a quick check that `ncvar` is a single character string (for example `"productivity"`), ensuring that the code references a single, well-defined variable from each data frame.

```{r}
# Build the common date set
common_dates <- sort(intersect(unique(modis_df$date_key), unique(noaa20_df$date_key)))

message("Found ", length(common_dates), " matching dates")
stopifnot(length(common_dates) > 0)   # fail fast if nothing matches

# ncvar should be a single string with the variable name, e.g. "productivity"
stopifnot(is.character(ncvar), length(ncvar) == 1)

```

## Calculate monthly pixel-wise ψ (VIIRS – MODIS)

For each month where both MODIS-Aqua and VIIRS-NOAA20 have data, we compute the unbiased relative difference ψ at every grid cell.

1.  We subset each dataset to a single month (`dt`) and keep only latitude, longitude, and the productivity variable.
2.  We join the MODIS and VIIRS slices on their shared `(latitude, longitude)` coordinates so that each row represents a single pixel observed by both sensors.
3.  We convert the productivity values to numeric (defensive step in case they arrive as character/factor).
4.  We compute ψ for each pixel using: $$
    \psi = \frac{\text{netPP}_{\text{VIIRS}} - \text{netPP}_{\text{MODIS}}}
                 {(\text{netPP}_{\text{VIIRS}} + \text{netPP}_{\text{MODIS}}) / 2}
    $$ where VIIRS (NOAA20) is the minuend and MODIS is the subtrahend.
5.  We reshape the ψ values into a 2-D matrix with rows ordered by latitude (south → north) and columns by longitude (west → east).
6.  For each month, we store the Unix timestamp, ψ matrix, and associated lat/lon grid in a list. This list becomes the basis for building the NetCDF file and subsequent timeseries and map visualizations.

```{r}
psi_results <- list()

for (kk in seq_along(common_dates)) {
  dt <- common_dates[kk]
  
  # Select coords + the variable, then rename value columns to keep them distinct
  s_slice <- modis_df  |>
    dplyr::filter(date_key == dt) |>
    dplyr::select(latitude, longitude, !!all_of(ncvar)) |>
    dplyr::rename(value_modis = !!sym(ncvar))
  
  n_slice <- noaa20_df |>
    dplyr::filter(date_key == dt) |>
    dplyr::select(latitude, longitude, !!all_of(ncvar)) |>
    dplyr::rename(value_noaa20 = !!sym(ncvar))
  
  # Intersection on (lat, lon)
  paired <- dplyr::inner_join(s_slice, n_slice, by = c("latitude", "longitude"))
  if (nrow(paired) == 0) {
    message("No overlapping cells on ", dt, "; skipping")
    next
  }
  
  # Ensure numeric (in case the field came in as character/factor)
  paired <- paired |>
    mutate(
      value_modis   = as.numeric(value_modis),
      value_noaa20 = as.numeric(value_noaa20)
    )
  
  # ψ = (A - B)/((A + B)/2); here SNPP is minuend, NOAA20 is subtrahend (swap if you prefer)
  paired <- paired |>
    mutate(psi = calculate_psi_dr(value_noaa20, value_modis, zero_denominator = "NA"))
  # If you want absolute relative difference: paired <- mutate(paired, psi = abs(psi))
  
  # Build ψ matrix: rows = latitude (S→N), cols = longitude (W→E)
  lon_vals <- sort(unique(paired$longitude))
  lat_vals <- sort(unique(paired$latitude))
  psi_mat  <- matrix(NA_real_, nrow = length(lat_vals), ncol = length(lon_vals))
  
  r_idx <- match(paired$latitude,  lat_vals)
  c_idx <- match(paired$longitude, lon_vals)
  psi_mat[cbind(r_idx, c_idx)] <- paired$psi
  
  # Unix timestamp @ 16th 00:00:00 UTC (already normalized earlier)
  ts_utc <- as.numeric(as.POSIXct(dt, tz = "UTC"))
  
  psi_results[[length(psi_results) + 1L]] <- list(
    timestamp = ts_utc,
    psi       = psi_mat,
    lon       = lon_vals,
    lat       = lat_vals
  )
  
  message("Appended ψ for ", format(as.Date(dt, origin="1970-01-01"), "%Y-%m-%d"))
}

```

## Build a 3D ψ array for NetCDF output

At this stage, `psi_results` is a list where each element contains the ψ field for a single month:

-   `psi_results[[k]]$psi` – 2D matrix of ψ values on a lat–lon grid
-   `psi_results[[k]]$lat` – latitude vector for that month
-   `psi_results[[k]]$lon` – longitude vector for that month
-   `psi_results[[k]]$timestamp` – Unix time stamp for the monthly center

To write an ERDDAP-style NetCDF file, we first need to:

1.  Extract the **time axis** as a numeric vector (seconds since 1970-01-01).
2.  Confirm that all monthly ψ fields share the **same latitude and longitude grid** (so they can be stacked into a single cube).
3.  Assemble a 3D array `psi_array[time, lat, lon]` by stacking the monthly ψ matrices along the time dimension.

This 3D array, together with `time_secs`, `lat_vals`, and `lon_vals`, forms the core data structure that we pass into the NetCDF writer.

```{r}
n_time <- length(psi_results)
stopifnot(n_time > 0L)

time_secs <- vapply(
  psi_results,
  function(x) as.numeric(x$timestamp),
  numeric(1)
)


# Take grid directly from first psi_results entry
lat_vals <- psi_results[[1]]$lat
lon_vals <- psi_results[[1]]$lon

n_lat <- length(lat_vals)
n_lon <- length(lon_vals)

psi_array <- array(NA_real_, dim = c(n_time, n_lat, n_lon))

for (ti in seq_len(n_time)) {
  slab     <- psi_results[[ti]]$psi   # matrix [lat, lon]
  slab_lat <- psi_results[[ti]]$lat
  slab_lon <- psi_results[[ti]]$lon
  
  # Ensure grid is consistent across all time slices
  if (!identical(slab_lat, lat_vals) || !identical(slab_lon, lon_vals)) {
    stop("psi_results[[", ti, "]] lat/lon grid does not match the first slice.")
  }
  if (!identical(dim(slab), c(n_lat, n_lon))) {
    stop("psi_results[[", ti, "]]$psi has dim ",
         paste(dim(slab), collapse = "x"),
         " but expected ", n_lat, "x", n_lon)
  }
  
  # Directly stack on [time, lat, lon]
  psi_array[ti, , ] <- slab
}

```

## Save results to a NetCDF file

Now that we have a complete 3D ψ cube (`psi_array[time, lat, lon]`) and the corresponding coordinate vectors, we write the results to disk as a CF/ERDDAP-style NetCDF file. The helper function `write_psi_netcdf()`:

-   Validates that the array and coordinate dimensions are consistent.
-   Derives the latitude/longitude bounds from `lat_vals` and `lon_vals`.
-   Builds a descriptive filename that encodes the sensors and spatial subset.
-   Creates a NetCDF file with `time`, `latitude`, `longitude`, and `psi` variables in ERDDAP-friendly order.
-   Attaches global metadata describing the project, region, and geospatial coverage.

The returned `out_path` points to the NetCDF file on disk, which we can then use for plotting, sharing, or publishing to ERDDAP.

```{r}
out_path <- write_psi_netcdf(
  psi_array   = psi_array,          # [time, lat, lon]
  time_secs   = time_secs,          # numeric seconds since 1970-01-01
  lat_vals    = lat_vals,
  lon_vals    = lon_vals,
  sensor1     = sensor1,            # "noaa20"
  sensor2     = sensor2,            # "modis"
  out_dir     = ODATA_DIR,
  region_label = "West Coast subset",
  overwrite   = TRUE                # or FALSE if you want safety
)

```

## Read ψ NetCDF and reconstruct the time axis

To generate summary statistics and visualizations, we first reopen the NetCDF file we just wrote and pull out the core pieces we need:

-   3D ψ array on the longitude, latitude, time grid
-   1D longitude and latitude coordinate vectors
-   Raw time coordinate

The `time` variable is stored as an offset from 1970-01-01T00:00:00Z (either in days or seconds, depending on the writer). We convert this numeric time axis back into calendar dates so that we can group ψ by month and build timeseries plots.

```{r}
psi_nc <- nc_open(out_path)

# psi is stored as [lon, lat, time] based on your ncvar_def
psi_arr  <- ncvar_get(psi_nc, "psi")        # 3D array
lon_vals <- ncvar_get(psi_nc, "longitude")  # 1D
lat_vals <- ncvar_get(psi_nc, "latitude")   # 1D
time_raw <- ncvar_get(psi_nc, "time")       # seconds since 1970-01-01T00:00:00Z

nc_close(psi_nc)

# Heuristic: treat values < 1e6 as days since 1970, otherwise seconds
if (max(time_raw, na.rm = TRUE) < 1e6) {
  # time_raw is in days since 1970-01-01
  date_vec <- as.Date(time_raw, origin = "1970-01-01")
} else {
  # time_raw is in seconds since 1970-01-01
  date_vec <- as.Date(
    as.POSIXct(time_raw, origin = "1970-01-01", tz = "UTC")
  )
}

```

## Convert the ψ cube to a monthly mean time series

The NetCDF file stores ψ as a full 3D field over longitude, latitude, and time. To build a 1D time series of spatially averaged ψ, we:

1.  Expand the lon–lat–time grid into a long-format data frame where each row represents a single pixel at a single month.
2.  Flatten the 3D ψ array into a vector and attach it to this grid in the same order.
3.  Attach the corresponding calendar date for each time index.
4.  Group by year and month and compute the **mean ψ across all pixels** for each month, producing a single ψ value per month.
5.  Construct a plotting date (using the 16th of each month) so the time series can be easily visualized and compared to other monthly products.

```{r}
# Expand grid over lon, lat, and *index* of time, then attach date
ntime <- length(date_vec)

psi_df <- expand.grid(
  longitude = lon_vals,
  latitude  = lat_vals,
  t_index   = seq_len(ntime)
)

# Flatten psi_arr in a matching order: [lon, lat, time]
psi_df$psi <- as.vector(psi_arr)

# Attach the Date for each t_index
psi_df$date <- date_vec[psi_df$t_index]

# At this point, psi_df$date should *not* be 1970-01-01 everywhere:
# head(psi_df$date)

monthly_data <- psi_df %>%
  mutate(
    year  = year(date),
    month = month(date, label = TRUE)
  ) %>%
  group_by(year, month) %>%
  summarise(
    monthly_mean = mean(psi, na.rm = TRUE),
    .groups = "drop"
  )

# Convert month abbrev → numeric
monthly_data$month_num <- match(monthly_data$month, month.abb)

# Build a plotting date (16th of each month)
monthly_data$date <- as.Date(
  sprintf("%04d-%02d-16", as.integer(monthly_data$year), monthly_data$month_num)
)

```

## Visualize time series of spatially averaged ψ

Finally, we visualize the monthly mean ψ values as a time series. Each point represents the **spatial average of ψ over the West Coast subset** for a given month, so values near zero indicate good agreement between VIIRS and MODIS netPP, while larger positive or negative values indicate months where one sensor consistently reports higher or lower productivity than the other.

We also add horizontal dashed lines at ±0.05 to highlight a ±5% difference band. Points that fall within this band correspond to months where the two products generally agree within about 5%, whereas values outside this range indicate larger differences.

```{r}
ggplot(monthly_data, aes(x = date, y = monthly_mean)) +
  geom_point(color = "black", size = 2) +
  geom_line(color = "black", linewidth = 1) +
  geom_hline(yintercept = 0.05,  linetype = "dashed", color = "red") +
  geom_hline(yintercept = -0.05, linetype = "dashed", color = "blue") +
  scale_x_date(
    date_breaks = "3 month",        # tick every month
    date_labels = "%b %Y"           # e.g., "Jan 2018"
  ) +
  labs(
    title = expression(bar(psi)[month] * " Timeseries: Unbiased Relative Difference VIIRS - MODIS (West Coast)"),
    x = "Month",
    y = "Unbiased Relative Difference (Psi)"
  ) +
  ylim(-0.1, 0.1) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # tilt labels so they fit
  )


```

## Visualize the spatial variability of mean ψ

In addition to the time series, it is useful to see **where** VIIRS and MODIS tend to differ the most. Here we collapse the full ψ cube across time and compute the **long-term mean ψ at each grid cell**:

-   Positive values (red) indicate locations where VIIRS netPP is, on average, higher than MODIS.
-   Negative values (blue) indicate locations where VIIRS netPP is, on average, lower than MODIS.
-   Values near zero (white) correspond to areas of good long-term agreement.

We then plot this mean ψ field over the West Coast subset. The color scale is clipped to the range $$-0.2, 0.2$$ to highlight differences within ±20%, and the coastline is added for geographic context.

```{r}
mean_map_df <- psi_df %>%
  group_by(longitude, latitude) %>%
  summarise(
    psi_mean = mean(psi, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(mean_map_df, aes(x = longitude, y = latitude, fill = psi_mean)) +
  geom_raster() +
  borders("world",
          xlim = c(-135, -115),
          ylim = c(30, 50),
          colour = "black",
          size = 0.4) +
  coord_quickmap(xlim = c(-135, -115), ylim = c(30, 50)) +
  scale_fill_gradient2(
    name = expression(bar(psi)),
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-0.2, 0.2),
    na.value = "grey80"
  ) +
  labs(
    title = expression("Spatial Variability " * bar(psi) * " (VIIRS – MODIS), West Coast"),
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()

```
